From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-07 Pi+/paper/discussion.tex
done with intro draft for now
Add more text in discussion

@@

\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
symmetric monoidal categories, strong symmetric monoidal functors, and symmetric monoidal natural transformations, to
the 2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, and its construction is known
in the literature~\cite{abramskyAbstractScalarsLoops2005}. Concretely, the objects of $\FSM[\CCat{C}]$ are given by
lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$. An morphism between $(n,A)$ and
$(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and
$\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction
$\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning
each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on
$C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$. The free
symmetric monoidal category has been used to study concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri
nets~\cite{baezCategoriesNets2021}, combinatorial structures~\cite{fioreCartesianClosedBicategory2008}, quantum
mechanics~\cite{abramskyAbstractScalarsLoops2005}, bicategorical models of (differential) linear
logic~\cite{melliesTemplateGamesDifferential2019}. In HoTT, the free symmetric monoidal groupoid $\FSM[A]$ over a
groupoid $A$ can be given by $\dsum{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs.
These HITs and the proof of their universal property have been considered
in~\citet*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.

\paragraph{Curry-Howard-Lambek correspondence} In~\citet{curryCurryEssaysCombinatory1980}, Lambek extended the
Curry-Howard correspondence to cartesian-closed categories. In this work, we have established a correspondence between a
fragment of the $\PiLang$ family of languages and symmetric monoidal groupoids.

Univalent Subuniverses, Univalent Typoids.

Higher order rewriting theory in constructive type theory.

Free monoids, free commutative monoids, normalisation of monoids.

@@

Computational group theory (in HoTT?)

Presentation of the free symmetric monoidal groupoid on one generator, proof of Mac Lane's coherence theorem in HoTT.

Alternatively, we can describe them using HITs, prove they have the universal property and hence equivalent.
The full Pi is FSMG 1, and the normalised fragment of Pi is M1.

>>>>

Pi types -- Natural number -- Finite sets
1-combinator -- Generators of Sn -- 1-paths
2-combinator -- Relations of Sn -- 2-paths

In HoTT, univalence characterises the path type in the universe as equivalences of types. The map $\term{idtoeqv} : A
\id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term $\term{ua} : A \eqv B \to A \id_{\UU}
B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its
extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p == \term{ua}(\term{idtoeqv(p)})$ are generally added as
postulates when formalising in Agda. Together, $\term{ua}$ and $\term{ua}-\beta$ give the full univalence axiom $(A \eqv
B) \eqv (A \id_{\UU} B)$.

% Let's think of the $\PiLang$ combinators as describing the inhabitants of the dentity type of finite types.

By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse. 
% By the property of being univalent, this subuniverse gives a model of the univalence axiom. 
The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
The $\term{ua}$ map corresponds to synthesing a program from an equivalence (which, in general, is of course
undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
sound, while $\term{ua-\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
this suggests a new computational interpretation of the univalence principle, which provides an intution on why certain
constructions are hard (or impossible in general case).

We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
doesn't have one, but we could show externally that it satisfies univalence.

@@

\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
\to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories was
studied in~\cite{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a groupoid
and the proof of its universal property has been considered in~\cite{piceghelloCoherenceMonoidalGroupoids2020}.

Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
$A$ can be given by $\dsum{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
and the proof of their universal property have been considered
in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.

The proof of the universal property of the $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric
monoidal categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite
set. The existence of the proof is folklore, and we have produced a new proof of it while working in constructive type
theory.

\paragraph{Curry-Howard-Lambek correspondence} In~\citet{curryCurryEssaysCombinatory1980}, Lambek extended the
Curry-Howard correspondence to cartesian-closed categories. In this work, we have established a correspondence between a
fragment of the $\PiLang$ family of reversible programming languages and symmetric monoidal groupoids. The Curry-Howard
part of this correspondence with reversible logic was established in~\cite{sparksSuperstructuralReversibleLogic2014}.

\paragraph{Rewriting} We presented a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.

There exists an algorithm, due to~\citet{knuthSimpleWordProblems1970}, that, when succeeds, constructs a well-behaved
rewriting system for an arbitrary finite set of (undirected) equations. It did not work for us, producing too many
equations, and proving correctness and termination was intractable.

We chose to encode permutations as adjacent transpositions corresponding to words in $\Sn$, since that closely
corresponds to the syntax of $\PiLang$ combinators. Other representations of permutations would give a different
denotational semantics for $\PiLang$ combinators. Permutations can be encoded as listed vectors or matrices, inductively
generated trees (Motzkin trees), Young diagrams, or String diagrams, but the difficulty of formalising them in type
theory varies depending on the encoding.

Rewriting systems and word problems have a long history of being formalised in proof assistants. In the recent past,
higher order rewriting systems have been formalised in proof assistants like {homotopy.io} and Lean and Coq. The use of
HoTT to study rewriting has been considered in~\cite{krausCoherenceWellFoundednessTaming2020}.

\paragraph{Computational group theory} Coxeter relations are used in computational group theory to study XXX problems.

\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
subuniverses.

@@

Formalised proofs of Mac Lane's coherence theorem.
